---
title: "CS-201 First Look at R"
author: "Amber Camp"
date: 8/26/26
format: html
editor: visual
toc: true
---

This is the first class session walking through code together!

We will be programming in R and in this class will primarily use `qmd` documents to keep things tidy and presentable. Each code chunk below is where you will run R code. Feel free to add notes as we go along.

## Navigating RStudio

By default, everyone should have the same four-pane layout:

1.  **Top left**: Source editor

2.  **Bottom left**: Console (and Terminal and Background Jobs)

3.  **Top Right**: Environment

4.  **Bottom Right**: Files, Plots, Help, etc.

### Customizing look & feel

Feel free to customize your RStudio.\
Go to Tools \> Global Options \> Appearance to select a color scheme you like.

I also personally like to enable these:

- Tools \> Global Options \> Code \> Insert matching parens/quotes

- Tools \> Global Options \> Console \> Show syntax highlighting in console input

I don't recommend rearranging your pane layout at this time.

## Running Code

Try running some code

### Load Library

```{r}
library(ggplot2)
```

### Some scripting

```{r}

quiz1 <- 88
quiz2 <- 94
quiz3 <- 79
# assignment operator to assign values using left facing arrow, are now variables and named values

quiz1
# gives quiz1 value 

# averaging
(quiz1 + quiz2 +quiz3) / 3

# mean
#mean(quiz1 + quiz2 + quiz 3) #not correct, just adds everything up
#mean(quiz1, quiz2, quiz 3) #also wrong, only returns the first average
mean(c(quiz1, quiz2, quiz3)) #correct

list <-c(quiz1, quiz2, quiz3)

quiz3 <-99

average <- mean(list)
average
round(average, 2)

```

```{r}
###List of More Scores

scores <- c(88, 94, 79, 91, 85)
scores
#c for combine

mean(scores)
min(scores)
max(scores)
length(scores)

summary(scores)

students <- c("a","b", "c", "d", "e")
students

#bar plot
barplot(scores, names.arg = students)

barplot(scores, names.arg = students, ylim = c(0, 100), main = "TITLE",
xlab = "pupils", y = "what they got")
#c can catonate, make a list
```

\

## Introducing `ggplot2`

#### Make a dataframe

```{r}

grades_df <- data.frame(student = students, score = scores)
grades_df

summary(grades_df) # of rows
nrow(grades_df) # of cols
ncol(grades_df) # of dimension
dim(grades_df)



```

#### Try ggplot

```{r}
ggplot(data = grades_df, aes(x = student, y = score)) # gives a blank canvas graph 

ggplot(data = grades_df, aes(x = reorder( student, -score), y = score, group = 1)) + 
  geom_col(fill = "steelblue") + 
  geom_point() +
  geom_line() +
  theme_minimal() + 
  labs(title = "BEST GRAPH EVER",
        subtitle = "some quiz scores",
        x = "da students",
        y = "what they scored on their quiz")
```

#### Mutate the dataframe

```{r}
#adding a variable tidyverse way
library(tidyverse)

#command shift m for this character the pipe operator
grades_df <- grades_df %>% 
  mutate(study_hours = c(2, 5, 1, 4, 3))
  
```

#### Make a scatterplot

```{r}
ggplot(data = grades_df, aes(x = study_hours, y = score)) + 
  geom_point(color = "purple") +
  geom_smooth(method = "lm", se = FALSE) +
  geom_line (color = "pink")


# cannot use fill, use color instead, don't use reorder

```

## Extra Credit (1pt)

What result do you expect when you run this expression?

```{r}
1 + 1 + 1
quiz1 + quiz2
1 + 1 + 1 - 3
0.1 + 0.1 + 0.1 - 0.3
options(scipen = 999)

print(0.1, 20)
print(0.3, 20)
print(0.5, 20)
```

What did you observe?

> Answer:

Why do you think this is happening? (You can look online for help)

> Answer:
